\sys: Scalable, Property-Guided\\ Program Regeneration

### 2025-07-31 docs/plos/draft.tex
Many changes
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\newcommand{\todo}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape Kv{\textalpha}sir}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\texttt{#1}\xspace}
\newcommand{\xxx}{\colorbox{red!30}{xxx}\xspace}
\newcommand{\rtype}[1]{\setlength{\fboxsep}{1.75pt}\colorbox{lightperiwinkle}{\texttt{#1}}}
\newcommand{\promptbox}[1]{%
  \vspace{0.5em}%
  \hspace*{-0.0135\columnwidth}%
  \fbox{%
    \parbox[t]{0.95\columnwidth}{\tt #1}%
  }%
  \vspace{0.5em} % adjust this value as desired
}

@@

\title{\sys: Controllable Program Regeneration}
\author{Evangelos Lamprou}

@@

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are common-place, but as brittle and time-consuming as ever.
  Towards their automation, large language models (LLMs) have shown promise in automating such tasks,
  but their outputs are shaped by surface cues, lack
  semantic guarantees, and their usage often boil down to brittle prompting strategies.
  This paper presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis. Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a minimal set of verifiable properties, such as
  the program's input/output behavior, execution traces, or interaction with external databases.
  These properties are extracted and serve the dual purpose of guiding synthesis and guardrailing synthesized outputs.
  \sys unifies established techniques and inspires new ones. % TODO: Rework this
  A preliminary evaluation shows that \sys succeeds even on tasks where state-of-the-art LLMs fail.
\end{abstract}

@@

\section{Introduction}
% \begin{figure}[t]
%   % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
%   \includegraphics[width=.9\columnwidth]{figs/kvasir_overview.pdf}
%   \caption{\textbf{\sys overview}
% Given a transformation query, and a source program, \sys extracts a minimal set
%   of properties,
%   guided by a logic engine and knowledge base.
%   It then synthesizes a new program $P'$ that satisfies the query, verifying
%   that wanted properties are present and unwanted properties are absent in $P'$.
% }
%   \label{fig:overview}
% \end{figure}

@@

This paper explores the feasibility of combining declarative property
constraints with LLM-based synthesis to support \emph{property-guided program regeneration}.
The key idea is to treat the original program as an untrusted reference from
which to extract minimal, verifiable properties (\eg I/O behavior,
function signatures, program traces), and to synthesize a new implementation that satisfies those
constraints without overfitting to potentially spurious details of the original source code.
The engine for this experimentation is the \sys prototype.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties.
At a high level, given a source program and a query, \sys uses the provided
logical constraints and its internal knowledge base to extract a set of
feasible and relevant program properties.
It then prompts an LLM to synthesize
candidate programs, which are subsequently verified against the desired properties.
This approach treats program transformation as projecting the original program over key properties.
Intuitively, regeneration then leverages LLM's high-dimensional internal representation~\cite{jin2024emergent,tao2024llms,huh2024platonicrepresentationhypothesis}, 
which lifts the projection back to a concrete implementation~(\cref{fig:projection}).
Unlike prompt-based approaches that require
extensive curation of examples~\cite{dilhara2024unprecedented, khattab2023dspy, cummins2024donttransformcodecode},
\sys treats the input program as an
untrusted artifact and explicitly projects only minimal, verifiable properties
into the regeneration process.

@@

\heading{Rethinking program transformations}
XXX Here, talk about bidirectional analysis, how now having a program that extracts 
a propoerty means that plugging it into sys will allow you to regenerate programs that hold or max/min-inimize over that property

\heading{Deployment scenarios and limitations}
\sys supports a range of deployment scenarios.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating
third-party modules that must be adapted to local conventions or hardened
against security vulnerabilities.
% For example, a developer can regenerate a C library as idiomatic
% JavaScript while preserving I/O equivalence, or decompose a monolithic function
% into modular components suitable for sandboxing.
% However, \sys is not designed for whole-program synthesis or large-scale migrations.
% Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module---where transformation goals can be explicitly stated and verified in isolation.
% To this end, \sys can function not only as a standalone tool but also as a
% library, where other components can invoke it as a programmable,
% constraint-guided regeneration module.

@@

@@

Consider a complex application, like the one
sketched in \cref{fig:ex-large-app}.
A single codebase includes many needs for transformations.
For example, a developer going through older parts of the codebase written by 
a previous team might realize that many functions are susceptible to buffer-overflow 
either through malice or by accident.
Then, thinking about a recently incorporated scripting component 
that allows players to make custom extensions to key parts of client-side game logic using JavaScript 
gives the idea tha maybe code that deals with user input should all be moved inside 
the sandboxed script engine.
Furthermore, the development team would like to identify functions that could 
be broken-down further to assist with performance profiling.

\heading{The state of the art and practise}
Currently, a developer has access to a myriad of different approaches 
to attack the above problems, but all come with shortcomings.
Code quality tools~\cite{codeql, MORE} and specialized transformation systems might be able to detect 
code-smells or certain classes of vulnerabilities but are often very highly 
tied to their target lagnuage.
As the programming-language landscape is becoming increasingly diverse,
and developers are building systems where multiple different languages are runtimes co-exist,
these limitations cannot be ignored.
Program synthesis could be an attractive avenue that allows 
the developer to 
The ever-present limitation of traditional program synthesis techniques is their 
inability to scale beyond narrowly-defined computational domains.
This scalability barrier is what statistical techniques like LLMs 
have promised to solve.
An LLM could in principle be tasked with performing all the afformentioned trasnformations.
However, their problems X, Y, Z. 
There have been systems that provide ranging amounts of control of LLM-generated 
programs, but there is a large gap of a unifying framework that identifies 
key parts of the problems and applies LLMs in a way that underlines their strenghths 
and allows the programmer to steer their output and be aware of incorrect generations.

@@

\section{Example}
\label{sec:example}

@@

This examples illustrates the capabilities of \sys by applying it to a representative game
application.
This example is selected to highlight the
diversity of transformation goals that arise even within a single codebase:
security hardening, cross-language migration, modernization of unsafe idioms,
and structural refactoring.

@@

\begin{wrapfigure}[4]{r}{.53\columnwidth}
\begin{minted}{prolog}
fsignt(p_,S):-fsignt(p,S).
graph(p_,I,O):-graph(p,I,O),
               crash(p,I).
:-crash(p_,I),graph(p_,I,_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \texttt{P\_} to exhibit the same
I/O behavior on representative inputs that would cause a buffer overflow in the original program $P$.
but not on the regenerated program $P^\prime$.
\sys first arrives the set of properties to extract from the
original program, given the query:
$\{\ttt{input(P, I)}, \ttt{graph(P, I, O)}, \ttt{signature(P, S)}\}$, where
\texttt{input(P, I)} means that \sys will use extract a set of inputs from the original program
by prompting an LLM with the program's source code and asking it to generate a set of inputs that
would cause a buffer overflow in the original program;
\texttt{graph(P, I, O)} extracts the client-observable I/O behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \texttt{signature(P, S)} extracts each function's signature using a language-aware parser.

After 30.4 seconds and two attempts, \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Args(int argc, char **argv) {
static char args[MAX_C];
int i; size_t remaining = MAX_C - 1;
args[0] = 0;
for (i = 1; i < argc; i++) {
  strncat(args, argv[i], remaining);
  remaining = MAX_C - strlen(args) - 1;
  if (i != argc - 1 && remaining > 0) {
    strncat(args, " ", remaining);
    remaining = MAX_C - strlen(args) - 1;
  } }
  return args; }
\end{minted}

@@

\begin{wrapfigure}[5]{r}{.46\columnwidth}
\begin{minted}{prolog}
language(p_, javascript).
graph(p_, Ijs, Ojs) :-
     graph(p, Ic, Oc),
     tr(Ic, Ijs),
     tr(Oc, Ijs).
\end{minted}
\end{wrapfigure}
The query specifies that the regenerated program \texttt{P\_} must be
implemented in JavaScript and must preserve observable behavior over a
representative set of inputs and outputs.
To bridge differences in
representation between C and JavaScript, the query uses \texttt{llm\_translate}
to convert each input and output into equivalent forms suitable for the target language.
\sys first determines the minimal set of properties to extract from the
original function: $\{\texttt{graph(P, I, O)}\}$.
It then invokes \texttt{llm\_translate} to produce
corresponding JavaScript representations of each example input and output, such
that the translated function will reproduce identical string formatting
behavior.
After 29.4 seconds, having generated 20 translated input-output pairs, \sys synthesizes the following JavaScript function:

\begin{minted}{javascript}
function cmdArgs(argv) {
    return argv.slice(1).join(" ");
}
\end{minted}
Notice that the regenerated program has a slightly different signature than the original, 
since JavaScript arrays include built-in information about their length,
which makes the \ttt{argc} argument obsolete.
The synthesis backend had the freedom to make this change since the query did not specify a signature property.

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[2]{r}{.56\columnwidth}
\begin{minted}{prolog}
graph(p_,I,O):-graph(p,I,O).
#min N: len(p_, N).
\end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{graph(p)}
from the original program
the captures captures the input-output behavior on a generated set of
inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

After 15.2 seconds and two iterations, \sys synthesizes the following idiomatic and portable version:
\begin{minted}{c}
#include <math.h>
float Q_rsqrt(float number) {
    return 1.0f / sqrtf(number); }
\end{minted}
The regenerated implementation is easier to verify, portable across compilers, and avoids undefined behavior.

@@

Given the query to the right, \sys generates a program that consists of two
functions whose composition is equivalent to the original implementation and
that preserve both I/O behavior and SQL trace properties (where a trace records
the sequence of queries executed against the database).

@@

The resulting program which now conists of two functions is:
\begin{minted}{javascript}
function connectDB() {
  const sqlite = require('sqlite');
  return sqlite.open({filename: 'game.db'}); }

@@

\heading{Key results}
\sys is able to effectively transform all programs across the above tasks,
producing high-quality outputs that satisfy user-specified constraints.
A total
of six property extraction/verification plugins (\ttt{graph}, \ttt{fsignt}, \ttt{sql\_trc}, and \ttt{llm\_translate})
were required to perform the above transformations, each
contributing either a set of extracted and verified properties, or a
transformation, and a set of pre- and post-conditions that the logic engine
uses to construct the regeneration plan.

@@

\begin{wrapfigure}{r}{.4\columnwidth}
  % https://docs.google.com/drawings/d/1d0piUrtExfTtUh_RuPfJFt1uZFOBEp0k-mHKSnogu-4/edit
  \includegraphics[width=.4\columnwidth]{figs/kvasir_levels.pdf}
  \label{fig:levels}
  \caption{The four key ingredients 
  of \sys that can be developed in parallel.}
\end{wrapfigure}
For \sys to enable developers
to express transformation goals declaratively---as constraints over the desired
properties of regenerated code---and automatically synthesizes implementations
that satisfy those constraints.

@@

\heading{Extensibility through plugins}
In order to maximize the applicability of the \sys framework 
towards many different kinds of transformations,
it is designed as an extensible platform rather than a monolithic tool.
All core functionality is provided by plugins, each contributing
the following functionality in the form three runtime hooks:
(1) \ttt{apply} which obtains static or dynamic properties from the original program (\eg
I/O traces, type signatures, or side-effect summaries),
(2) \ttt{verify} that will check the regenerated program against these properties (re-extracting them and comparing them to the original ones),
and
(3) \ttt{know} adds pre/post-conditions that inform regeneration planning.
The vision of \sys it to encourages community-driven growth, where specialized plugins---\eg for
security invariants or performance contracts---can coexist and interoperate
within a single regeneration framework~(\cref{sec:discussion}).

\heading{A declarative query language}
To detach itself from the details of the underlying analysis and

@@

\heading{Regeneration guardrails}
The final ingredient is a synthesis–verification loop that turns declarative goals into trustworthy code.
Given extracted properties and a query, Kvasir synthesizes candidate implementations using LLMs, leveraging their flexibility and knowledge of diverse programming idioms.
Each candidate is then checked against the declared properties: does it preserve the specified behavior, meet language or security constraints, and satisfy any additional invariants?
Failures trigger retries, refined prompts, or additional property extraction,
while successes provide an auditable proof of meeting the user’s intent,
at least as far as the properties allow (which is still a lot more than most LLM-based systems can guarantee).

\heading{Prototype implementation}
The \sys prototype implemetation uses Python vXXX,
the Pluggy library to define pluggable hooks, compiles queries to Clingo and uses the clasp
solver to arrive at a stable model which is converted to synthesis plans. 
XXX plugins have been implemented. All plugins range from XXX to XXX lines-of-code,
with the entire codebase being XXX lines-of-code.

\heading{A benchmark suite for program regeneration}
For research on program regeneration to flourish, it needs a common benchmark
suite that captures the diversity and difficulty of real-world transformations.
Today, there is no equivalent of SPEC for performance or Defects4J for bug
fixing—researchers rely on ad hoc examples that are hard to compare and often
oversimplified. This lack of a shared baseline obscures progress: when one
approach claims success on “language migration” and another on “security
hardening,” we have no way to tell whether they solve the same problem, let
alone which approach generalizes better.

Kvasir takes a step in this direction with its prototype benchmarks.
This suite contains tasks across several dimensions: its
ability to regenerate correct and maintainable
code, the impact of property-driven guidance,
and practical considerations around applying it
to diverse inputs. This evaluation is
preliminary and intended to illustrate
feasibility and trade-offs.

% TODO: Describe what a good benchmark suite would look like
The benchmark suite comprises 37 tasks spanning simple translation, obfuscation
removal, and security hardening (\cref{tab:benchmarks}). Inputs were drawn
from Rosetta Code, popular npm and PyPi packages, known supply-chain attacks,
and selected obfuscated programs from the IOCCC.

@@

In microbenchmarks, \sys generated idiomatic rewrites (e.g., replacing unsafe
inverse square root implementations with standard library calls) while
maintaining observable behavior.

>>>>

\newcommand{\todo}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape Kv{\textalpha}sir}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\texttt{#1}\xspace}
\newcommand{\xxx}{\colorbox{red!30}{xxx}\xspace}
\definecolor{lightperiwinkle}{rgb}{0.85, 0.97, 1.0}
\newcommand{\prop}[1]{\setlength{\fboxsep}{1.75pt}\colorbox{lightperiwinkle}{\texttt{#1}}}
\newcommand{\promptbox}[1]{%
  \vspace{0.5em}%
  \hspace*{-0.0135\columnwidth}%
  \fbox{%
    \parbox[t]{0.95\columnwidth}{\tt #1}%
  }%
  \vspace{0.5em} % adjust this value as desired
}

@@

\title{\sys: Scalable, Property-Guided\\ Program Regeneration}
\author{Evangelos Lamprou}

@@

\begin{abstract}
  Software evolution tasks---like security hardening, porting critical components to safer 
  languages, or refactoring opaque legacy code---are commonplace, but as brittle and time-consuming as ever.
  Towards their automation, large language models (LLMs) show promise.
  However, their outputs are shaped by surface cues, lack semantic guarantees, and often rely on brittle prompting strategies.
  This paper presents \sys, a system for controllable program regeneration that
  combines declarative property constraints with LLM-guided synthesis.
  Unlike
  prior approaches, \sys treats the input program as an untrusted artifact and
  explicitly projects it into a set of verifiable properties, such as
  the program's input/output behavior, execution traces, or interaction with external data stores.
  These properties are extracted and serve the dual purpose of guiding synthesis and guardrailing synthesized outputs.
  \sys unifies established techniques and inspires new ones. % TODO: Rework this
  A preliminary evaluation shows that \sys succeeds even on tasks where state-of-the-art LLMs fail.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
% \begin{figure}[t]
%   % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
%   \includegraphics[width=.9\columnwidth]{figs/kvasir_overview.pdf}
%   \caption{\textbf{\sys overview}
% Given a transformation query, and a source program, \sys extracts a minimal set
%   of properties,
%   guided by a logic engine and knowledge base.
%   It then synthesizes a new program $P'$ that satisfies the query, verifying
%   that wanted properties are present and unwanted properties are absent in $P'$.
% }
%   \label{fig:overview}
% \end{figure}

@@

This paper explores the feasibility of combining declarative property
constraints with LLM-based synthesis to support \emph{property-guided program regeneration}.
The key idea is to treat the original program as an untrusted reference from
which to extract minimal, verifiable properties (\eg input-output behavior,
function signatures, program traces), and to synthesize a new implementation that satisfies those
constraints without overfitting to potentially spurious details of the original source code.
The engine for this experimentation is the \sys prototype.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties.
At a high level, given a source program and a query, \sys uses the provided
logical constraints and its internal knowledge base to extract a set of
feasible and relevant program properties.
It then prompts an LLM to synthesize
candidate programs, which are subsequently verified against the desired properties.
This approach treats program transformation as projecting the original program over key properties.
Intuitively, regeneration then leverages LLM's high-dimensional internal representation~\cite{jin2024emergent,tao2024llms,huh2024platonicrepresentationhypothesis}, 
which lifts the projection back to a concrete implementation~(\cref{fig:projection}).
% Unlike prompt-based approaches that require
% extensive curation of examples~\cite{dilhara2024unprecedented, khattab2023dspy, cummins2024donttransformcodecode},
% \sys treats the input program as an
% untrusted artifact and explicitly projects only minimal, verifiable properties
% into the regeneration process.

@@

\heading{Rethinking program transformations}
% Here, talk about bidirectional analysis, how now having a program that extracts 
% a propoerty means that plugging it into sys will allow you to regenerate programs that hold or max/min-inimize over that property
Traditional transformation frameworks typically apply a one-way mapping: 
a program is analyzed to detect issues or features, and, separately, 
a hand-written rule or heuristic is applied to transform it. 
\sys enables a fundamentally different, \emph{bidirectional} view. 
Once an analysis can \emph{extract} a property,
that same analysis, when plugged into \sys's declarative query language and 
verification loop, can be used to \emph{regenerate} programs that 
satisfy, minimize, or maximize that property. 
This shifts program analyses from passive diagnostics into \emph{active drivers} 
of change: any property extractor becomes, effectively, a transformation oracle. 
The result is an extensible ecosystem where existing analyses not only tell us 
what is wrong or suboptimal, but also allow us to automatically create 
semantics-preserving, property-optimized alternatives.

\heading{Deployment scenarios}
\sys supports a range of deployment scenarios.
It is particularly well-suited for use during maintenance and auditing phases, where developers seek to refactor or replace components, especially when the original code is difficult to understand, modify, or is unavailable and the component exists only as a binary.
It can also assist during development, for instance when integrating
third-party modules that must be adapted to local conventions or hardened
against security vulnerabilities.
% For example, a developer can regenerate a C library as idiomatic
% JavaScript while preserving input-output equivalence, or decompose a monolithic function
% into modular components suitable for sandboxing.
% However, \sys is not designed for whole-program synthesis or large-scale migrations.
% Its strength lies in scoped regeneration tasks, typically at the granularity of a function, method, or small module---where transformation goals can be explicitly stated and verified in isolation.
% To this end, \sys can function not only as a standalone tool but also as a
% library, where other components can invoke it as a programmable,
% constraint-guided regeneration module.

@@

\heading{Availability}
The \sys early-stage prototype in available as an open-source MIT-licensed artifact at:
\begin{center}
  \url{https://github.com/blinded}
\end{center}

@@

Consider a complex application, like the one
sketched in \cref{fig:ex-large-app}.
A single codebase includes many needs for transformations.
For example, a developer going through older parts of the codebase written by 
a previous team might realize that many functions are susceptible to buffer-overflow attacks.
Then, thinking about a recently incorporated JavaScript scripting component 
that allows players to make custom extensions to key parts of client-side game logic, 
suggests that perhaps code that deals with user input should all be moved inside 
the sandboxed runtime.
Furthermore, the development team would like to identify functions that could 
be broken-down further to assist with performance profiling.

\heading{The state of the art and practise}
Currently, a developer has access to a myriad of different approaches 
to attack the above problems, but all come with shortcomings.
Code quality tools~\cite{avgustinov2016ql, Fowler99,Mens04,Myers16} and specialized transformation systems might be able to detect 
code-smells or certain classes of vulnerabilities but are often very highly 
tied to their target language.
As the programming-language landscape is becoming increasingly diverse,
and developers are building systems where multiple different languages are runtimes co-exist,
these limitations cannot be ignored.
Program synthesis could be an attractive avenue that allows 
the developer to 
The ever-present limitation of traditional program synthesis techniques~\cite{jha2010oracle, raza2018disjunctive, singh2016blinkfill,wu2023programming, harp:ccs:2021, yaghmazadeh2018automated} is their 
inability to scale beyond narrowly-defined computational domains.
This scalability barrier is what statistical techniques like LLMs 
have promised to solve.
An LLM could in principle be tasked with performing all the aforementioned transformations.
However, they come with no shortage of well-studied shortcomings~(\cref{sec:intro}).
There have been systems that provide ranging amounts of control of LLM-generated 
programs~\cite{huang2023anplnaturalprogramminginteractive,li2025mccoderstreamliningmotioncontrol,barua2025pygencollaborativehumanaiapproach, tihanyi2025new},
but there is a large gap of a unifying framework that identifies 
key parts of the problems and applies LLMs in a way that underlines their strengths,
end-users to steer their output, and---more importantly---be aware when 
their outputs are \emph{wrong}.

@@

\section{Examples}
\label{sec:example}

@@

This example is used to highlight the
diversity of transformation goals that arise even within a single codebase:
security hardening, cross-language migration, modernization of unsafe idioms,
and structural refactoring.

@@

\begin{wrapfigure}[4]{r}{.53\columnwidth}
\begin{minted}{prolog}
fsignt(p_,S):-fsignt(p,S).
graph(p_,I,O):-graph(p,I,O),
               crash(p,I).
:-crash(p_,I),graph(p_,I,_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P\_} to exhibit the same
input-output behavior on representative inputs that would cause a buffer overflow in the original program \ttt{P}.
but not on the regenerated program \ttt{P\_}.
\sys first arrives the set of properties to extract from the
original program, given the query:
\{\prop{input(P, I)}, \prop{graph(P, I, O)}, \prop{fsignt(P, S)}\}, where
\prop{input(P, I)} means that \sys will use extract a set of inputs from the original program
by prompting an LLM with the program's source code and asking it to generate a set of inputs that
would cause a buffer overflow in the original program;
\prop{graph(P, I, O)} extracts the client-observable input-output behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \prop{fsignt(P, S)} extracts each function's signature using a language-aware parser.
After 30.4 seconds and two attempts, \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Args(int argc, char **argv) {
static char args[MAX_C];
int i; size_t remaining = MAX_C - 1;
args[0] = 0;
for (i = 1; i < argc; i++) {
  strncat(args, argv[i], remaining);
  remaining = MAX_C - strlen(args) - 1;
  if (i != argc - 1 && remaining > 0) {
    strncat(args, " ", remaining);
    remaining = MAX_C - strlen(args) - 1;
  } }
  return args; }
\end{minted}

@@

\begin{wrapfigure}[5]{r}{.43\columnwidth}
\begin{minted}{prolog}
language(p_, js).
graph(p_, Ijs, Ojs) :-
     graph(p, Ic, Oc),
     tr(Ic, Ijs),
     tr(Oc, Ijs).
\end{minted}
\end{wrapfigure}
The query specifies that the regenerated program \texttt{P\_} must be
implemented in JavaScript and must preserve observable behavior over a
representative set of inputs and outputs.
To bridge differences in
representation between C and JavaScript, the query invokes the \ttt{tr} plugin
to convert each input and output into equivalent forms suitable for the target language.
\sys first determines the minimal set of properties to extract from the
original function: \{\prop{graph(P, I, O)}\}.
It then, again, invokes \ttt{tr} to produce
corresponding JavaScript representations of each example input and output, such
that the translated function will reproduce identical string formatting
behavior.
After 29.4 seconds, having generated 20 translated input-output pairs, \sys synthesizes the following JavaScript function:

\begin{minted}{javascript}
function cmdArgs(argv) {
    return argv.slice(1).join(" ");
}
\end{minted}
Notice that the regenerated program has a slightly different signature than the original, 
since JavaScript arrays include built-in information about their length,
which makes the \ttt{argc} argument obsolete.
The synthesis backend had the freedom to make this change since the query did not require preserving
the function signature.

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[2]{r}{.6\columnwidth}
\begin{minted}{prolog}
graph(p_,I,O):-graph(p,I,O).
#min N: len(p_, N).
\end{minted}
\end{wrapfigure}
\sys extracts the set of properties \{\prop{graph(P, I, O)}, \prop{len(P, N)}\}
from the original program
the captures captures the input-output behavior on a generated set of
inputs.
In this case, where the goal is to optimize an aspect of the program---its length---\sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes no difference after two consecutive regenerations.

After 15.2 seconds and three iterations, \sys synthesizes the following idiomatic and portable version:
\begin{minted}{c}
#include <math.h>
float Q_rsqrt(float number) {
    return 1.0f / sqrtf(number); }
\end{minted}
The regenerated implementation is easier to verify, portable across compilers, and avoids undefined behavior.

@@

Given the query to the right, \sys generates a program that consists of two
functions whose composition is equivalent to the original implementation and
that preserve both input-output behavior and SQL trace properties (where a trace records
the sequence of queries executed against the database).

@@

The resulting program which now consists of two functions is:
\begin{minted}{javascript}
function connectDB() {
  const sqlite = require('sqlite');
  return sqlite.open({filename: 'game.db'}); }

@@

\heading{Key result}
\sys is able to effectively transform all programs across the above tasks,
producing high-quality outputs that satisfy user-specified constraints.
A total
of seven property extraction/verification plugins (\ttt{graph}, \ttt{fsignt}, \ttt{crash}, \ttt{sql\_trc}, and \ttt{tr})
were required to perform the above transformations, each
contributing either a set of extracted and verified properties, or a
transformation, and a set of pre- and post-conditions that the logic engine
uses to construct the regeneration plan.

@@

\begin{wrapfigure}{r}{.4\columnwidth}
  % https://docs.google.com/drawings/d/1d0piUrtExfTtUh_RuPfJFt1uZFOBEp0k-mHKSnogu-4/edit
  \includegraphics[width=.4\columnwidth]{figs/kvasir_levels.pdf}
  \label{fig:levels}
  \caption{The four key ingredients 
  of \sys that can be developed in parallel.}
\end{wrapfigure}
The insights that fuel \sys's design~(\cref{sec:intro}) lead to the following key ingredients.

@@

\heading{Extensibility through plugins}
In order to maximize the applicability of the \sys framework 
towards many different kinds of transformations,
it is designed as an extensible platform rather than a monolithic tool.
All core functionality is provided by plugins, each contributing
the following functionality in the form three runtime hooks:
(1) \ttt{apply} which obtains static or dynamic properties from the original program (\eg
input-output traces, type signatures, or side-effect summaries),
(2) \ttt{verify} that will check the regenerated program against these properties (re-extracting them and comparing them to the original ones),
and
(3) \ttt{know} adds pre/post-conditions that inform regeneration planning.
The vision of \sys it to encourage community-driven growth, where specialized plugins---\eg for
security invariants or performance contracts---can coexist and interoperate
within a single regeneration framework~(\cref{sec:discussion}).

\heading{A declarative query language}
The dizzying rate of progress in the space of 
code-generating LLMs calls for an interface 
that abstracts over the specific strengths and weaknesses of specific models 
and allows the underlying implementation to advance separately from the rest of the system.
An obvious choice: a high-level, declarative language that 
allows for reasoning irrespective of the underlying synthesis backend.

The \sys query language allows one to describe transformation intent as logical constraints over program 
properties. 
These queries are resolved against a \emph{knowledge base} of extracted 
program properties and plugin-provided pre- and post-conditions that specify when an analysis is applicable, or not.
This represents the space of behaviors and invariants 
that must be preserved, eliminated or optimized.

@@

\heading{Regeneration guardrails}
Synthesis leverages an LLM as an 
(imperfect but flexible) \emph{bidirectional analysis}, capable of 
mapping between representations.

To make this work, the final ingredient is a synthesis and verification loop that turns declarative goals into trustworthy code.
Given extracted properties and a query, \sys synthesizes candidate implementations using LLMs, leveraging their flexibility and knowledge of diverse programming idioms.
Each candidate is then checked against the declared properties: does it preserve the specified behavior, meet language or security constraints, and satisfy any additional invariants?
Failures trigger retries, a refined prompt.
A success provides an auditable proof of meeting the user's intent,
at least as far as the extracted properties allow (which is still a lot more than most LLM-based systems can guarantee).

\heading{Prototype implementation}
The \sys prototype implemetation uses Python v3.13,
the \ttt{pluggy} system~\cite{pluggy} to define plugin hooks,
compiles queries to the Clingo language~\cite{DBLP:journals/corr/GebserKKS14} and uses the \ttt{clasp}~\cite{gebser2007clasp}
solver to arrive at a stable model which is converted to synthesis plans. 
A total of seven plugins have been implemented (\ttt{graph}, \ttt{tr}, \ttt{len}, \ttt{sql\_trc}, \ttt{call}, \ttt{crash}, \ttt{fsignt}).
Each plugin ranges from 9 to 51 lines-of-code,
with the entire codebase being 605 lines-of-code.

\heading{A benchmark suite for program regeneration}
For research on program regeneration to flourish, it needs a common benchmark
suite that captures the diversity and difficulty of real-world transformations.
Today, there is no equivalent of SPEC for performance or Defects4J for bug
fixing---researchers and practioners rely on ad hoc examples that are hard to compare and often
oversimplified.
This lack of a shared baseline obscures progress. 
A useful set of benchmarks should include
programs that
(1) are real and non-trivial,
(2) are written in a variety of languages, 
(3) cover a wide range of computational domains.
Each benchmark should include an original program, 
the same program that undergoes the transformation described in natural-language 
in an accompanying file, and a set of tests that verify that the transformation was succesful.
Following the lead of benchmarking efforts~\cite{SPEC_CPU2017, dacapo2006, koala:atc:2025}, the suite should be made open-source, community 
first, and an established technical steering commitee should guide extensions and modifications towards maintaining the suite's relevance.

\sys takes a step in this direction with its prototype benchmarks.
This suite contains tasks across several dimensions: its
ability to regenerate correct and maintainable
code, the impact of property-driven guidance,
and practical considerations around applying it
to diverse inputs. This evaluation is
preliminary and intended to illustrate
feasibility and trade-offs.

The benchmark suite comprises 37 tasks spanning language translation, obfuscation
removal, and security hardening~(\cref{tab:benchmarks}).
Inputs were drawn from established sources
like the Rosetta Code project, popular npm and PyPi packages, high-profile supply-chain attacks,
and selected obfuscated programs from the IOCCC.

@@

In microbenchmarks, \sys generated idiomatic rewrites (\eg replacing unsafe
inverse square root implementations with standard library calls) while
maintaining observable behavior.
